---
title: "52542 Generalized Linear Models: Theory and Application - Final Paper"
author: "Omer Brandes - 302910476"
date: "March 4, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS)
library(boot)
```

##A brief description of the problem:

We study counts of rat sightings in the city of Madrid.
The brown rat lives with mankind and adversely affects public health by transmission
of diseases, bites and allergies. Better understanding behavioural and spatial corre-
lation aspects of this species can contribute to its effective management and control.
We explore weakly to moderately correlated covariates based on distances to broken
sewers, feeding grounds and markets as well as population density.
The data were collected in Madrid city.
It has a municipal surface area of about 605 km2 and 3.2 million inhabitants.
In the context of large urban settlements, approximately 3% of the households have
rats in their immediate environment, e.g.in compost heaps, gardens or unsecured
rubbish bins. In the city of Madrid, direct sightings of rats and/or
cockroaches or signs of their presence (e.g. droppings, burrows, gnaw marks, etc.) can
be reported by citizens to the Technical Unit for Vector Control (TUVC). Only reports
from people who declare to have sighted themselves any kind of these pest animals or
their vital sign(s) in areas falling within the administrative borders of the municipality
of Madrid are accepted. Records of the location and time of observation are entered in a dedicated database.
The data used in this study contain the locations and dates of 6693 validated rat
sightings reported to the TUVC from 1 January 2010 to 31 December 2013.
[^1]

[^1]: Gräler, Benedikt, Carlos Ayyad, and Jorge Mateu. "Modelling count data based on weakly dependent spatial covariates using a copula approach: application to rat sightings." Environmental and Ecological Statistics 24, no. 3 (2017): 433-448.


##Exploratory analysis of the data:
```{r, echo=FALSE}
library(readxl)
rats <- read_excel("rats.xlsx")
attach(rats)
vars = colnames(rats)
```

We have the following variables in the data set:   
`r vars`

We will not use the id in the analysis since it obviously can't help us predict the response variable. 
In addition, since 'total.count' is just the sum of 'rat.count' and 'ckr.count', and therefore doesn't add any new information, we will not use it in the analysis as well. 
Some of the data:   


Summary of the variables that will be used in the analysis:     
```{r echo=FALSE}
summary(rats[,c(3,4,7:9)])
```


###Exploring each variable:    

The red line in the plots represents the mean value.     

rat.count. We can see in the Histogram below that we have a large number of zeros, we will try to deal with this when we fit the models later on:   

```{r echo=FALSE}
par(mfrow=c(2,2))
hist(rat.count, breaks = 25)
plot(rat.count)
abline(h = mean(rat.count), col = 'red')
boxplot(rat.count)
```

ckr.count:    

```{r echo=FALSE}
par(mfrow=c(2,2))
hist(ckr.count, breaks = 18)
plot(ckr.count)
abline(h = mean(ckr.count), col = 'red')
boxplot(ckr.count)

```

market.dist:    

```{r echo=FALSE}
par(mfrow=c(1,2))
plot(market.dist)
abline(h = mean(market.dist), col = 'red')
boxplot(market.dist)
```

sewer.dist:   

```{r echo=FALSE}
par(mfrow=c(1,2))
plot(sewer.dist)
abline(h = mean(sewer.dist), col = 'red')
boxplot(sewer.dist)

```

catfeeding.dist:   

```{r echo=FALSE}
par(mfrow=c(1,2))
plot(catfeeding.dist)
abline(h = mean(catfeeding.dist), col = 'red')
boxplot(catfeeding.dist)

```


we have a few large observations in rat.count, we will try to see if these are outliers that should be
removed from the dataset. We have three rat.count observation larger than two time the standard deviation
of rat.count:
```{r}
bigCount = which(rat.count > mean(rat.count)+2*sd(rat.count))  

rats[bigCount,-c(2,5,6)]
```

We expect the above observation to have below average distance from markets and sewers, but we first need to figure out wether cat feeding stations encourage rats or not:

```{r echo=FALSE}
plot(catfeeding.dist,rat.count)
```

We can see that the negative trend in the above plot suggests that cat feeding stations encourage rats. 
Therefore, we expect that the possible outliers will also have below average cat feeding distance.  

The possible outliers are marked with full red dots, the red horizontal line represents the mean value of the variable:

```{r echo=FALSE}
par(mfrow = c(2,2))
plot(market.dist)
points(c(20,21,43) ,market.dist[c(20,21,43)], col = 'red', pch = 16)
abline(h = mean(market.dist), col = 'red')

plot(sewer.dist)
points(c(20,21,43) ,sewer.dist[c(20,21,43)], col = 'red', pch = 16)
abline(h = mean(sewer.dist), col = 'red')

plot(catfeeding.dist)
points(c(20,21,43) ,catfeeding.dist[c(20,21,43)], col = 'red', pch = 16)
abline(h = mean(catfeeding.dist), col = 'red')
```

And we can see that the outliers are below average on all three variables, and therefore seems like we should
keep them in the dataset.   


plotting some graphs for further visualisation:

```{r echo=FALSE}
par(mfcol = c(2,2))
plot(catfeeding.dist,rat.count)
plot(market.dist,rat.count)
plot(ckr.count,rat.count)
```


```{r echo=FALSE}
par(mfrow = c(1,1))
pairs(cbind(rats[vars[c(3,4,7:9)]])) 
```

Covariance matrix - most are very low, with some excpeitons like rat.count/ckr.count and total.count, but that is expected:
```{r echo=FALSE}
cor(rats[vars[1:9]])
```


  
##Model fitting:
We will start by fitting a Linear Model to the dataset. Since our response variable (rat.count) is a count variable, we don't expect the LM to fit well:
```{r}
reg1 = lm(rat.count ~ ckr.count + market.dist + sewer.dist + catfeeding.dist)
reg1_summary = summary(reg1)
reg1_summary
```

And indeed the model shows a very low R-squared value: `r reg1_summary$r.squared`.    

Next, we will try to fit a poisson model:

```{r}
glim1 = glm(rat.count ~ ckr.count + market.dist + sewer.dist + catfeeding.dist,      
            family = "poisson", data = rats)
summary(glim1)
```

We now attempt to add spatial variables to the model and see if there is a spatial trend, 
we fit a few models with different spatial variables such as x,y,x^2 ,y^2 , and compare their AIC and Devience:
```{r echo=FALSE}
glimSpatial1 = glm(rat.count ~ ckr.count + market.dist + sewer.dist + catfeeding.dist
                   +xc , family = "poisson", data = rats)
glimSpatial2 = glm(rat.count ~ ckr.count + market.dist + sewer.dist + catfeeding.dist
                   +yc , family = "poisson", data = rats)
glimSpatial3 = glm(rat.count ~ ckr.count + market.dist + sewer.dist + catfeeding.dist
                   +xc + yc , family = "poisson", data = rats)
glimSpatial4 = glm(rat.count ~ ckr.count + market.dist + sewer.dist + catfeeding.dist
                   +xc + yc + xc^2 + yc^2, family = "poisson", data = rats)
```
```{r}
cbind(glimSpatial1$aic, glimSpatial2$aic,glimSpatial3$aic,glimSpatial4$aic)
cbind(glimSpatial1$deviance, glimSpatial2$deviance,glimSpatial3$deviance,glimSpatial4$deviance)
```

And finally, we compare the models with and without the sptial data:
```{r}
cbind(glimSpatial1$aic, glim1$aic)
cbind(glimSpatial1$deviance, glim1$deviance)
```

We will evaluate the poisson model with the spatial data using the Goodnes of Fit test:
The GOF test indicates that the Poisson model doesn't fit the data (p < 0.05).
```{r}
1 - pchisq(summary(glimSpatial1)$deviance, summary(glimSpatial1)$df.residual)
```

This is expected because it seems like we have over-dispersion - the variance is much larger than the Expected Value (using the mean as an estimate to the Expected Value):
```{r}
mean(rat.count)
var(rat.count)
```

In addition, we need to find a model that accounts for the zero inflation we noted earlier.   

Negative-Binomial model:
```{r}
glimNB = glm.nb(rat.count ~ ckr.count + market.dist + sewer.dist + catfeeding.dist, 
                data = rats)
```
this model leads to an AIC of glimNB$aic.
The devience is: `r glimNB$deviance`
The Residual Sum of Squares: `r sum((rat.count - glimNB$fitted.values)^2)`

```{r}
glimNB_spatial = glm.nb(rat.count ~ ckr.count + market.dist + sewer.dist + catfeeding.dist + xc, 
                        data = rats)
```
Adding the spatial data of the x coordinates yields AIC glimNB_spatial$aic.
Devience: glimNB_spatial$deviance.
The Residual Sum of Squares is better: `r sum((rat.count - glimNB_spatial$fitted.values)^2)`
   
So far NB model with spatial data gives the best aic and devience. 
```{r}
rbind((cbind(glimNB$aic,glimNB$deviance)),
      (cbind(glimNB_spatial$aic,glimNB_spatial$deviance )))
```

Goodness of fit for NB models.
The GOF test indicates that the Poisson model fits the data (p > 0.05).

```{r}
1 - pchisq(summary(glimNB_spatial)$deviance, 
           summary(glimNB_spatial)$df.residual)
```


Now we attempt to improve the model using a stepwise process:
```{r}
step(glimNB_spatial,direction="both")
```
Seems like we can remove the catfeeding.dist variable.



```{r}
glimNB_spatial_2 = glm.nb(rat.count ~ ckr.count + market.dist + sewer.dist + xc, 
                        data = rats)

```

Goodness of fit improved: `r 1 - pchisq(summary(glimNB_spatial_2)$deviance,summary(glimNB_spatial_2)$df.residual)`
  
AIC: `r glimNB_spatial_2$aic`
Devience: `r glimNB_spatial_2$deviance`
Residual sum of squares: `r sum((rat.count - glimNB_spatial_2$fitted.values)^2)`

   
    
In conclusion, we fitted the Negative-Binomial model to the data, and improved it using a stepwise process.
The graph below shows the response variable (black empty dots) and the fitted valued (red full dots):

```{r}
par(mfrow = c(1,1))
plot(rat.count)
points(glimNB_spatial_2$fitted.values,col='red', pch = 19)
```




